### 2017-02-28 2017mergingmerges/section/04-Fusion.tex
prose S4 S5

\eject{}
% -------------------------------------
Fig Fig.~\ref{fig:Fusion:Def:StepPair} defines function \ti{tryStepPair} which decides which of two processes to advance.

@@

Clause (SharedPull) applies when the left process wants to pull from a shared channel that the other process either pulls from or pushes to. We know that there is already a value in the stream buffer variable, because the state for that channel is $@pending@_S$. The result instruction copies the value from the stream buffer variable into variable specific to the left source process, and the corresponding $@have_S@$ channel state in the result label records that it has done so.

>>>>

\eject{}
% -------------------------------------
Fig.~\ref{fig:Fusion:Def:StepPair} defines function \ti{tryStepPair} which decides which of two processes to advance.

@@

Clause (SharedPull) applies when the left process wants to pull from a shared channel that the other process either pulls from or pushes to. We know that there is already a value in the stream buffer variable, because the state for that channel is $@pending@_S$. The result instruction copies the value from the stream buffer variable into a variable specific to the left source process, and the corresponding $@have@_S$ channel state in the result label records that it has done so.

### 2017-02-28 2017mergingmerges/section/05-Evaluation.tex
notes
Half way though cleaning S4
prose S4 S5

% -----------------------------------------------------------------------------
\section{Fusion in Practice}
\label{s:Evaluation}

@@

% -----------------------------------------------------------------------------
\subsection{Fusability}
\label{s:FusionOrder}
When we fuse a pair of processes we commit to a particular interleaving of instructions from each process. When we have at least three processes to fuse, the choice of which two to handle first can determine whether this fused result can then be fused with the third process. Consider our @alternatives@ example from \S\ref{s:EvaluationOrder}, here it is again:
\begin{code}
  alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
  alternates sInA sInB sInC
   = let  s1   = alt2 sInA sInB
          s2   = alt2 sInB sInC
          sOut = zip s1 s2
     in   sOut
\end{code}

In our current system, if fuse the two @alt2@ processes together first, then try to fuse this result process with the downstream @zip@ process, then this final fusion transform fails. This happens because the first fusion transform commits to a sequential instruction interleaving where two output values \emph{must} be pushed to stream @s1@ first, before then pushing values to @s2@. As we discussed in \S\ref{s:EvaluationOrder}, the downstream @zip@ process needs to pull single values from @s1@ and @s2@ alternatively.

Dynamically, if we were to try to execute the result process and the downstream @zip@ process concurrently, then the execution would deadlock. Statically, when we try to fuse the result process with the downstrem @zip@ process then the deadlock is discovered and fusion fails. Deadlock happens when neither process can advance to the next instruction, and in the fusion algorithm this will manifest as the failure of the $tryStepPair$ function from Fig.\ref{fig:Fusion:Def:StepPair}. The $tryStepPair$ function determines which of the instructions from either process can be executed next, and when execution is deadlocked there are none.

On the upside, fusion failure is easy to detect. It is also easy to provide a report to the client programmer that describes why two particular processes could not be fused. The report is phrased in terms of the process definitions visible to the client programmer, instead of partially fused intermediate code. The joint labels used in the fusion algorithm represent which states each of the original processes would be in during a concurrent execution, and we provide the corresponding instructions as well as the abstract states of all the input channels. This reporting ability is \emph{significantly better} than that of prior fusion systems such as Repa~\cite{lippmeier2012:guiding}, as well as the co-recursive stream fusion of \cite{coutts2007stream}, and may other systems based on general purpose program transformations. In such systems it is usually not clear whether the fusion transformation even succeeded, and debugging why it might not have succeeded involves spelunking\footnote{def. spelunking: Exploration of caves, especially as a hobby. Usually not a science.} through many pages (sometimes hundreds of pages) of compiler intermediate representations.

In practice, the likelihood of fusion suceeding depends on the particular dataflow network being used, as well as the form of the processes in that network. For fusion of pipelines of standard combinators such as @map@, @fold@, @filter@, @scan@ and so on, fusion always succeeds. The process implementations of each of these combinators only pull one element at a time from their source streams, before pushing the result to the output stream, so there is no possiblity of deadlock. Deadlock can only happen when multiple streams fan-in to a process with multiple inputs, such as with @merge@. When the dataflow network has a single output stream then we use the method of starting from the process closes to the output stream, walking to towards the input streams, and fusing in successive processes as they occur. This allows the interleaving of the intermediate fused process to be dominated by the consumers, rather than producers, as consumers are more likely to have multiple input channels which need to be synchronized. In the worst case the fallback approach is to try all possible orderings of processes to fuse, assuming the client programmer is willing to wait for the search to complete.

@@

As with fusion transformation system, we must be careful that the size of the result code does not become too large when more and more processes are fused together. The left of Fig.\ref{fig:bench:outputsize} shows the maximum number of output states in the result when a particular number of processes are fused together in a pipelined-manner. To produce this graph we programmatically generated dataflow networks for \emph{all possible} pipelined combinations of the @map@, @filter@, @scan@, @group@ and @merge@ combinators, and tried all possible fusion orders consiting of adjacent pairs of processes. The @merge@ combinator itself has two inputs, so only works at the very start of the pipeline --- we present result for pipelines with and without a @merge@ at the start. The right of Fig.\ref{fig:bench:outputsize} shows the number of states in the result when the various combinations of combinators are fused in parallel, for example, we might have a @map@ and a @filter@ processing the same input stream. In both cases the number of states in the result process grows linearly with the number of processes. In all combinations, with up to 7 processes there are less than 100 states in the result process.

@@

On the other hand, Fig.\ref{fig:bench:exponential} shows the results for a pathological case where the size of the output program is exponential in the number if input processes. The source dataflow networks consists of N merge processes, N+1 input streams, and a single output stream. The output of each merge process is the input of the next, forming a chain of merges. In source notation the network for N = 3 is @sOut = merge sIn1 (merge sIn2 (merge sIn3 sIn4))@.

When fusing two processes the fusion algorithm essentially compares every state in the first process with every state in the second, computing a cross product. During the fusion transform, as states in the result process are generated they are added to a finite map --- the @instrs@ field of the process definition. The use of the finite map ensures that identical states are always combined, but genuinly different states always make it into the result.

In the worst case fusion of two processes produces O($n*m$) different states, where $n$ and $m$ are the number of states in each. If we assume the two processes have about the same number of states then this is O($n^2$). Fusing the next process into this result yields O($n^3$), so overall the worst case number of states in the result will be O($n^k$), where $k$ is the number of processes fused.

@@

Minimising the number of states in an intermediate process has the follow-on effect that the final fused result also has lower number of states. Provided we do not change the order of instructions that do not require synchronization with other processes, meaning @pull@ @push@ or @drop@, the fusability of the overall process network will not be affected.

>>>>

\eject{}
% -----------------------------------------------------------------------------
\section{Fusion in Practice}
\label{s:Evaluation}

@@

% -----------------------------------------------------------------------------
\subsection{Fusibility}
\label{s:FusionOrder}
When we fuse a pair of processes we commit to a particular interleaving of instructions from each process. When we have at least three processes to fuse, the choice of which two to handle first can determine whether this fused result can then be fused with the third process. Consider our @alternatives@ example from \S\ref{s:EvaluationOrder}, here it is again:
\begin{code}
  alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
  alternates sInA sInB sInC
   = let  s1   = alt2 sInA sInB
          s2   = alt2 sInB sInC
          sOut = zip s1 s2
     in   sOut
\end{code}

In our current system, if we fuse the two @alt2@ processes together first, then try to fuse this result process with the downstream @zip@ process, then this final fusion transform fails. This happens because the first fusion transform commits to a sequential instruction interleaving where two output values \emph{must} be pushed to stream @s1@ first, before then pushing values to @s2@. As we discussed in \S\ref{s:EvaluationOrder}, the downstream @zip@ process needs to pull single values from @s1@ and @s2@ alternatively.

Dynamically, if we were to try to execute the result process and the downstream @zip@ process concurrently, then the execution would deadlock. Statically, when we try to fuse the result process with the downstream @zip@ process then the deadlock is discovered and fusion fails. Deadlock happens when neither process can advance to the next instruction, and in the fusion algorithm this will manifest as the failure of the $tryStepPair$ function from Fig.\ref{fig:Fusion:Def:StepPair}. The $tryStepPair$ function determines which of the instructions from either process can be executed next, and when execution is deadlocked there are none.

On the upside, fusion failure is easy to detect. It is also easy to provide a report to the client programmer that describes why two particular processes could not be fused. The report is phrased in terms of the process definitions visible to the client programmer, instead of partially fused intermediate code. The joint labels used in the fusion algorithm represent which states each of the original processes would be in during a concurrent execution, and we provide the corresponding instructions as well as the abstract states of all the input channels. This reporting ability is \emph{significantly better} than that of prior fusion systems such as Repa~\cite{lippmeier2012:guiding}, as well as the co-recursive stream fusion of \cite{coutts2007stream}, and many other systems based on general purpose program transformations. In such systems it is usually not clear whether the fusion transformation even succeeded, and debugging why it might not have succeeded involves spelunking\footnote{def. spelunking: Exploration of caves, especially as a hobby. Usually not a science.} through many pages (sometimes hundreds of pages) of compiler intermediate representations.

In practice, the likelihood of fusion suceeding depends on the particular dataflow network being used, as well as the form of the processes in that network. For fusion of pipelines of standard combinators such as @map@, @fold@, @filter@, @scan@ and so on, fusion always succeeds. The process implementations of each of these combinators only pull one element at a time from their source streams, before pushing the result to the output stream, so there is no possiblity of deadlock. Deadlock can only happen when multiple streams fan-in to a process with multiple inputs, such as with @merge@. When the dataflow network has a single output stream then we use the method of starting from the process closest to the output stream, walking to towards the input streams, and fusing in successive processes as they occur. This allows the interleaving of the intermediate fused process to be dominated by the consumers, rather than producers, as consumers are more likely to have multiple input channels which need to be synchronized. In the worst case the fallback approach is to try all possible orderings of processes to fuse, assuming the client programmer is willing to wait for the search to complete.

@@

As with any fusion system, we must be careful that the size of the result code does not become too large when more and more processes are fused together. The left of Fig.~\ref{fig:bench:outputsize} shows the maximum number of output states in the result when a particular number of processes are fused together in a pipelined-manner. To produce this graph we programmatically generated dataflow networks for \emph{all possible} pipelined combinations of the @map@, @filter@, @scan@, @group@ and @merge@ combinators, and tried all possible fusion orders consiting of adjacent pairs of processes. The @merge@ combinator itself has two inputs, so only works at the very start of the pipeline --- we present result for pipelines with and without a @merge@ at the start. The right of Fig.~\ref{fig:bench:outputsize} shows the number of states in the result when the various combinations of combinators are fused in parallel, for example, we might have a @map@ and a @filter@ processing the same input stream. In both cases the number of states in the result process grows linearly with the number of processes. In all combinations, with up to 7 processes there are less than 100 states in the result process. 
%%% AR: Note that the split version also has only one merge?

@@

On the other hand, Fig.~\ref{fig:bench:exponential} shows the results for a pathological case where the size of the output program is exponential in the number of input processes. The source dataflow networks consists of N merge processes, N+1 input streams, and a single output stream. The output of each merge process is the input of the next, forming a chain of merges. In source notation the network for N = 3 is @sOut = merge sIn1 (merge sIn2 (merge sIn3 sIn4))@.

When fusing two processes the fusion algorithm essentially compares every state in the first process with every state in the second, computing a cross product. During the fusion transform, as states in the result process are generated they are added to a finite map --- the @instrs@ field of the process definition. The use of the finite map ensures that identical states are always combined, but genuinely different states always make it into the result.

In the worst case, fusion of two processes produces O($n*m$) different states, where $n$ and $m$ are the number of states in each. If we assume the two processes have about the same number of states then this is O($n^2$). Fusing the next process into this result yields O($n^3$), so overall the worst case number of states in the result will be O($n^k$), where $k$ is the number of processes fused.

@@

Minimising the number of states in an intermediate process has the follow-on effect that the final fused result also has lower number of states. Provided we do not change the order of instructions that do not require synchronization with other processes, meaning @pull@ @push@ or @drop@, the fusibility of the overall process network will not be affected.
